A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-09-10 doc/popl2013/response.txt

We'd first like to thank the reviewers for their detailed and useful feedback.
Rather than addressing each reviewer's questions separately, we have organized
our response based on the areas the questions covered.

First, there were multiple requests for more detail on the code example, the
use of colorings to describe partitions, and the algorithm used to perform 
efficient dynamic disjointness checks.  Each of these was covered in significant
detail in [1], {which we understand the reviewers may not have had access to.}

As described in [1], Legion does not attempt to solve the (in many cases still-open)
problem of partitioning data for any given application.  The body of the "color_circuit"
task would likely consist of a call to an external library (e.g. PARMETIS for graph-based
data structures).  The coloring is merely a way to capture the result of that external
computation for use by the Legion runtime.  The "build_lists" task is not strictly required.
While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
each element of the parent region, by building a list up front, the cost of those "downregion"
checks can be amortized over multiple iterations of the simulation.

In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.

Legion supports the "partition then allocate" model in which new elements allocated in a subregion
are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
any other regions.  The use of a non-empty initial coloring in the operational semantics simplifies
this in Core Legion by allowing extra unnamed locations to exist in every region to allow these
allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
can of worms?}

As several reviewers noted, the Legion type system is similar in many ways to existing region-based
typed systems, and this is by design.  Additions to the "standard" region-based type system were 
done just to capture the new features of Legion: colorings as a description of an arbitrary partitioning
of a region, a hierarchy of regions in which a location may be contained in multiple regions, the 
ability to capture the relationship between regions, and the ability to relax the determinism of 
execution via coherence modes.  Although these changes were made as contained as
possible, a proof that soundness was preserved by these additions seemed prudent.

The use of coherence modes in Legion differs from existing usages, so we tried to reduce confusion by
not reusing existing terminology.  While "exclusive" coherence corresponds to apparently-sequential
execution, "atomic" corresponds to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism, an important difference is that these coherence modes apply
locally in Legion, i.e. only between subtasks of the same parent task.  All three of these modes have
their uses in real-world applications - to exclude any of them would limit the applicability of the Legion
programming model.  We felt the additional complexity in the type system and operational semantics was
justified by the results that both soundness of the type system and ability to make local scheduling decisions
were noy compromised by the inclusion of these coherence modes.

Finally, the choice of large-step operational semantics was made to match the hierachical and distributed
scheduling used by the Legion runtime.  When coherence modes other than exclusive are used, there may be
multiple tasks modifying the same location of memory - the clobber set is the set of all such locations for
a given task.  The clobber set is used to bound the allowed non-determinism within a given task, and the
fact that the top-level task (which has no sibling tasks) has an empty clobber set guarantees that any 
execution of the program is consistent with a standard sequentially consistent memory model.

An advantage of the large-step semantics with the use of clobber sets is that our soundness result holds for
any scheduling of tasks that satisfies the "valid_interleave" rules.  We considered the use of small-step semantics
that explicitly track the state of every parallel task's execution, but the complexity of the constraints used
to decide which task(s) were able to take a step at any given point were onerous and would limit our soundness
guarantees to only that scheduling algorithm.

### 2012-09-10 doc/popl2013/response.txt

We'd first like to thank the reviewers for their detailed and useful feedback.
Rather than addressing each reviewer's questions separately, we have organized
our response based on the areas the questions covered.

First, there were multiple requests for more detail on the code example, the
use of colorings to describe partitions, and the algorithm used to perform 
efficient dynamic disjointness checks.  Each of these was covered in significant
detail in [1], which may have been unavailable to the reviewers previously but can
now be found at http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .

@@

As described in [1], Legion does not attempt to solve the (in many cases still-open)
problem of partitioning data for any given application.  The body of the "color_circuit"
task would likely consist of a call to an external library (e.g. PARMETIS for graph-based
data structures).  The coloring is merely a way to capture the result of that external
computation for use by the Legion runtime.  The "build_lists" task is not strictly required.
While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
each element of the parent region, by building a list up front, the cost of those "downregion"
checks can be amortized over multiple iterations of the simulation.

In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.

[Legion supports the "partition then allocate" model in which new elements allocated in a subregion
are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
any other regions.  The use of a non-empty initial coloring in the operational semantics simplifies
this in Core Legion by allowing extra unnamed locations to exist in every region to allow these
allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
can of worms?}]

As several reviewers noted, the Legion type system is similar in many ways to existing region-based
typed systems; this is by design.  Additions to the "standard" region-based type system were 
done to capture the new features of Legion: colorings as a description of an arbitrary partitioning
of a region, a hierarchy of regions in which a location may be contained in multiple regions, the 
ability to capture the relationship between regions, and the ability to relax the determinism of 
execution via coherence modes.  With these additional features a proof that soundness was preserved 
was a prerequisite for later proofs in our paper.

A description of the usage and intuition behind coherence modes was fully presented in [1] and
was omitted from this paper due to space constraints.
"Exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism.  However, unlike other implementations, coherence modes 
only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
to a portion of the task hierarchy.  This paper contains two new
results about coherence modes not contained in [1].  First, we give a formal semantics of the possible
interleavings allowed by 
different coherence modes and their interactions.  Second, we prove that coherence modes do not
impact the soundness of our type system or the correctness of our hierarchical scheduling algorithm.

### 2012-09-11 doc/popl2013/response.txt

In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.

[Legion supports the "partition then allocate" model in which new elements allocated in a subregion
are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
any other regions.  The use of a non-empty initial coloring in the operational semantics simplifies
this in Core Legion by allowing extra unnamed locations to exist in every region to allow these
allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
can of worms?}]

As several reviewers noted, the Legion type system is similar in many ways to existing region-based
typed systems; this is by design.  Additions to the "standard" region-based type system were 
done to capture the new features of Legion: colorings as a description of an arbitrary partitioning
of a region, a hierarchy of regions in which a location may be contained in multiple regions, the 
ability to capture the relationship between regions, and the ability to relax the determinism of 
execution via coherence modes.  With these additional features a proof that soundness was preserved 
was a prerequisite for later proofs in our paper.

A description of the usage and intuition behind coherence modes was fully presented in [1] and
was omitted from this paper due to space constraints.
"Exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism. However, unlike previous work, coherence modes 
only allow for reordering of subtasks with the same parent task.  This paper contains two useful
results about coherence modes.  First, we give a formal semantics of the possible interleavings allowed by 
different coherence modes and their interactions.  Second, we prove that coherence modes do not
impact the soundness of our type system or the correctness of our hierarchical scheduling algorithm.

### 2012-09-11 doc/popl2013/response.txt

We would first like to thank the reviewers for their detailed and useful feedback.
Rather than addressing each reviewer's questions separately, we have organized
our response based on the areas the questions covered.

@@

In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.

[Legion supports the "partition then allocate" model in which new elements allocated in a subregion
are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
any other regions.  The use of a non-empty initial coloring in the operational semantics simplifies
this in Core Legion by allowing extra unnamed locations to exist in every region to allow these
allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
can of worms?}]

As several reviewers noted, the Legion type system is similar in many ways to existing region-based
typed systems; this is by design.  Additions to the "standard" region-based type system were 
done to capture the new features of Legion: colorings as a description of an arbitrary partitioning
of a region, a hierarchy of regions in which a location may be contained in multiple regions, the 
ability to capture the relationship between regions, and the ability to relax the determinism of 
execution via coherence modes.  With these additional features a proof that soundness was preserved 
was a prerequisite for later proofs in our paper.

A description of the usage and intuition behind coherence modes was fully presented in [1] and
was omitted from this paper due to space constraints.
"Exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism.  However, unlike other implementations, coherence modes 
only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
to a portion of the task hierarchy.  This paper contains two new results about coherence modes 
not presented in [1].  First, we give a formal semantics of the possible interleavings allowed by 
different coherence modes and their interactions.  Second, we prove that coherence modes do not
impact the soundness of our type system or the correctness of our hierarchical scheduling algorithm.

### 2012-09-11 doc/popl2013/response.txt

We would first like to thank the reviewers for their detailed and useful feedback.
Rather than addressing each reviewer's questions separately, we have organized
our response based on the areas the questions covered.

First, there were multiple requests for more detail on the code example, the
use of colorings to describe partitions, and the algorithm used to perform 
efficient dynamic disjointness checks.  Each of these was covered in significant
detail in our prior work [1], which may have been unavailable to the reviewers previously but can
now be found at http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .

### 2012-09-11 doc/popl2013/response.txt

We thank the reviewers for their thoughtful, extensive, and very detailed comments.
Most of the comments are specifically about presentation; we acknowledge the issue,
and if the paper is accepted we will address the concerns raised.

Many of the reviewers raise similar questions, which we have grouped together in a
single response below in addition to answering issues raised by individual reviewers.
To keep the response to a reasonable length we have tried to address the most common
and significant issues.  In many cases, questions were asked which were addressed
in our prior published work [1], which may have been unavailable to the reviewers previously but can
now be found at http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .

Reviewers A, B, and C all asked for additional details about colorings.
In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.  As described in [1], Legion does not attempt to solve the (in many cases still-open)
problem of partitioning data for any given application.  In our example, the body of the "color_circuit"
task would require the programmer to compute a partition of the circuit graph and return a coloring
that assigns each circuit element a color corresponding to its piece of the partition.  
In production scientific codes this is often done by a call 
to an external library (e.g. PARMETIS for graph-based data structures).  
A coloring is merely a way to capture the result of the partitioning 
computation in a general format understood by the Legion runtime.

Reviewer A asked how "downregion" operations are performed and used.  The "downregion" operation
requires a dynamic check that the given location is valid in the provided logical region.  To
perform these checks the runtime remembers which locations have been allocated in each logical
region and dynamically checks that the location has been allocated for the given region.  The 
"build_lists" task is an example of an optimization done to reduce the number of "downregion" calls.
While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
each element of the parent region, by building a list typed on the subregion up front, the cost 
of those "downregion" checks can be amortized over multiple iterations of the simulation.

Both reviewers A and B inquired how the dynamic disjointness tests are performed.  This
is the primary subject of our prior work [1].

@@

[Legion supports the "partition then allocate" model in which new elements allocated in a subregion
are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
any other regions.  The use of a non-empty initial coloring in the operational semantics simplifies
this in Core Legion by allowing extra unnamed locations to exist in every region to allow these
allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
can of worms?}]

Reviewer B commented that the Legion type system is similar in several ways to existing region-based
typed systems.  While we reuse many features of prior region languages, we believe that Legion
represents a fundamentally new composition of these features.  Primarily Legion allows for first-class
logical regions which permit elements to exist in many regions simultaneously via multiple dynamic
partitions of regions.  Furthermore, the way the region information is leveraged both at compile-time 
and at runtime represent a new point in the design space of region languages.  While the proof of
soundness is not novel, we felt it necessary given the difference in semantics of logical regions 
from previous region languages.  The soundness proof was also a prerequisite for our later proofs
concerning coherence modes and hierarchical scheduling.

Both reviewers B and C thought coherence modes were an interesting technical contribution, but
would have liked a higher-level description to build intuition.
A description of the usage and intuition behind coherence modes was fully presented in [1] and
was omitted from this paper due to space constraints.  Our goal in this paper was to present
two new results about coherence modes not presented in [1].  First, we present a formal semantics 
of the possible interleavings allowed by different coherence modes and their interactions.  
Second, we prove that coherence modes do not impact the soundness of our type system or the 
correctness of our hierarchical scheduling algorithm.  A brief summary of coherence modes is:
"exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism.  Unlike other implementations, coherence modes 
only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
to a portion of the task hierarchy.

Reviewer D asked what is currently implemented.  Legion currently consists of a static type 
checker for programs written in Core-Legion and a C++ runtime library for fully-featured Legion programs.

Finally, reviewers B, C, and D expressed confusion about the nature of the clobber set.
The choice of large-step operational semantics was made to match the hierachical and distributed
scheduling used by the Legion runtime.  When coherence modes other than exclusive are used, there may be
multiple tasks modifying the same location of memory - the clobber set is the set of all such locations for
a given task.  The clobber set is used to bound the allowed non-determinism within a given task, and the
fact that the top-level task (which has no sibling tasks) has an empty clobber set guarantees that any 
execution of the program is consistent with a standard sequentially consistent memory model.

### 2012-09-11 doc/popl2013/response.txt

Reviewer A asked how "downregion" operations are performed and used.  The "downregion" operation
requires a dynamic check that the given location is valid in the provided logical region.  To
perform these checks the runtime remembers which locations have been allocated in each logical
region and dynamically checks that the location has been allocated for the given region.  The 
"build_lists" task is an example of an optimization done to reduce the number of "downregion" calls.
While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
each element of the parent region, by building a list typed on the subregion up front, the cost 
of those "downregion" checks can be amortized over multiple iterations of the simulation.

Both reviewers A and B inquired how the dynamic disjointness tests are performed.  This
is a primary subject of our prior work [1].

[Legion supports the "partition then allocate" model in which new elements allocated in a subregion
are guaranteed to exist in parent regions as well and preserve that subregion's disjointness with
any other regions.  The use of a non-empty initial coloring in the operational semantics simplifies
this in Core Legion by allowing extra unnamed locations to exist in every region to allow these
allocations without changing the sets of locations assigned to physical regions.  {Is this opening a
can of worms?}]

Reviewer B commented that the Legion type system is similar in several ways to existing region-based
typed systems.  While we reuse many features of prior region languages, we believe that Legion
represents a fundamentally new composition of these features.  Primarily Legion allows for first-class
logical regions which permit elements to exist in many regions simultaneously via multiple dynamic
partitions of regions.  Furthermore, the way the region information is leveraged both at compile-time 
and at runtime represent a new point in the design space of region languages.  While the proof of
soundness does not introduce novel proof techniques, it is also far from immediate that the type system
is sound and we feel the proof is necessary given large differences in the semantics of logical regions 
from previous region systems.  The soundness proof was also a prerequisite for our later proofs
concerning coherence modes and hierarchical scheduling.

Both reviewers B and C thought coherence modes were an interesting technical contribution, but
would have liked a higher-level description to build intuition.
A description of the usage and intuition behind coherence modes is presented in [1] and
was omitted from this paper due to space constraints.  Our goal in this paper was to present
two new results about coherence modes that are not in [1].  First, we present a formal semantics 
of the possible interleavings allowed by different coherence modes and their interactions.  
Second, we prove that coherence modes do not impact the soundness of our type system or the 
correctness of our hierarchical scheduling algorithm.  A brief summary of coherence modes is:
"exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism.  Unlike other implementations, coherence modes 
only allow for reordering of subtasks with the same parent task, permitting the containment of non-determinism
to a portion of the task hierarchy.

### 2012-09-12 doc/popl2013/response.txt

We thank the reviewers for their very detailed comments.  Most of the
comments are specifically about presentation; we acknowledge the
issue, and if the paper is accepted we will address the concerns
raised.

Many of the reviewers raise similar questions, which we have grouped
together in a single response below in addition to answering issues
raised by individual reviewers.  To keep the response to a reasonable
length we have tried to address the most common and significant
issues.  In many cases, questions were asked which were addressed in
our prior published work [1], which may have been unavailable to the
reviewers previously but can now be found at
http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .

Reviewers A, B, and C all asked for additional details about colorings.
In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.  As described in [1], Legion does not attempt to solve the (in many cases still-open)
problem of partitioning data for any given application.  In our example, the body of the "color_circuit"
task would require the programmer to compute a partition of the circuit graph and return a coloring
that assigns each circuit element a color corresponding to its piece of the partition.  The same
set of colors are used to identify the subregions that result from the partitioning.
In production scientific codes this is often done by a call 
to an external library (e.g. PARMETIS for graph-based data structures).  
A coloring is merely a way to capture the result of the partitioning 
computation in a general format understood by the Legion runtime.

The principle reason to support multiple partitions of the same data is to allow precise descriptions
of the sets of elements that will be accessed by a given task.  These sets are often different in 
different phases of an algorithm, and the increased precision is necessary to safely parallelize
applications such as the circuit example.  Although the subregions of different partitions are often
accessed with different privileges, this is a property of an algorithm rather than a Legion requirement.

@@

Reviewer A asked how "downregion" operations are performed and used.  The "downregion" operation
requires a dynamic check that the given location is valid in the provided logical region.  To
perform these checks the runtime remembers which locations have been allocated in each logical
region and dynamically checks that the location has been allocated for the given region.  The 
"build_lists" task is an example of an optimization done to reduce the number of "downregion" calls.
While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
each element of the parent region, by building a list typed on the subregion up front, the cost 
of those "downregion" checks can be amortized over multiple iterations of the simulation.

Both reviewers A and B inquired how the dynamic disjointness tests are performed.  This
is a primary subject of our prior work [1].

Reviewer B commented that the Legion type system is similar in several ways to existing region-based
typed systems.  While we reuse many features of prior region languages, we believe that Legion
represents a fundamentally new composition of these features.  Primarily Legion allows for first-class
logical regions which permit elements to exist in many regions simultaneously via multiple dynamic
partitions of regions.  Furthermore, the way the region information is leveraged both at compile-time 
and at runtime represent a new point in the design space of region languages.  While the proof of
soundness does not introduce novel proof techniques, it is also far from immediate that the type system
is sound and we feel the proof is necessary given large differences in the semantics of logical regions 
from previous region systems.  The soundness proof was also a prerequisite for our later proofs
concerning coherence modes and hierarchical scheduling.

Both reviewers B and C thought coherence modes were an interesting technical contribution, but
would have liked a higher-level description to build intuition.
A description of the usage and intuition behind coherence modes is presented in [1] and
was omitted from this paper due to space constraints.  Our goal in this paper was to present
two new results about coherence modes that are not in [1].  First, we present a formal semantics 
of the possible interleavings allowed by different coherence modes and their interactions.  
Second, we prove that coherence modes do not impact the soundness of our type system or the 
correctness of our hierarchical scheduling algorithm.  A brief summary of coherence modes is:
"exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism.  However, unlike other implementations, 
Legion's coherence modes only allow for reordering of subtasks with the same parent task, 
permitting the containment of non-determinism to a portion of the task hierarchy.

Reviewer A asked if the operational semantics would be simpler if the only relaxation of sequential
semantics came from an explicit "par(e1,e2)" construct.  While the semantics would indeed be simpler,
the use of a static construct (rather than one able to consider dynamic region mappings) would limit
the expressible forms of parallelism to a (well-studied) subset of what Legion programs can achieve.

Reviewer D asked what is currently implemented.  Legion currently consists of a static type 
checker for programs written in Core Legion and a C++ runtime library for fully-featured Legion programs.

Finally, reviewers B, C, and D expressed confusion about the nature of the clobber set.
The choice of large-step operational semantics was made to match the hierarchical and distributed
scheduling used by the Legion runtime.  When coherence modes other than exclusive are used, there may be
multiple tasks modifying the same location of memory - the clobber set is the set of all such locations for
a given task.  The clobber set is used to bound the allowed non-determinism within a given task, and the
fact that the top-level task (which has no sibling tasks) has an empty clobber set guarantees that any 
execution of the program is consistent with a standard sequentially consistent memory model.

An advantage of the large-step semantics with the use of clobber sets is that our soundness result holds for
any scheduling of tasks that satisfies the "valid_interleave" rules.  We considered the use of small-step semantics
that explicitly track the state of every parallel task's execution, but the complexity of the constraints used
to decide which task(s) were able to take a step at any given point were onerous and would limit our soundness
guarantees to only a particular scheduling algorithm.

### 2012-09-12 doc/popl2013/response.txt

We thank the reviewers for their very detailed comments.  Most of the
comments are specifically about presentation; we acknowledge the
issue, and if the paper is accepted we will address the concerns
raised.

Many of the reviewers raised similar questions, which we have grouped
together in a single response below in addition to answering issues
raised by individual reviewers.  To keep the response to a reasonable
length we have tried to address the most common and significant
issues.  In many cases, questions were asked which were addressed in
our prior published work [1], which may have been unavailable to the
reviewers previously but can now be found at
http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .

Reviewers A, B, and C all asked for additional details about colorings which were covered in [1].
In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.  As described in [1], Legion does not attempt to solve the (in many cases still-open)
problem of partitioning data for any given application.  In our example, the body of the "color_circuit"
task would require the programmer to compute a partition of the circuit graph and return a coloring
that assigns each circuit element a color corresponding to its piece of the partition.  The same
set of colors are used to identify the subregions that result from the partitioning.
In production scientific codes partitioning is often done by a call 
to an external library (e.g. PARMETIS for graph-based data structures).  
A coloring is merely a way to capture the result of the partitioning 
computation in a general format understood by the Legion runtime.

Reviewer C asked for clarification concerning the the relationship between multiple partitions and privileges.
The principle reason to support multiple partitions of the same data is to allow precise descriptions
of the sets of elements that will be accessed by a given task.  These sets are often different in 
different phases of an algorithm, and the increased precision is necessary to safely parallelize
applications such as the circuit example.  Although the subregions of different partitions are often
accessed with different privileges, this is a property of an algorithm rather than a Legion requirement.

@@

Reviewer A asked how "downregion" operations are performed and used.  The "downregion" operation
requires a dynamic check that the given location is valid in the provided logical region.  To
perform these checks the runtime remembers which locations have been allocated in each logical
region and dynamically checks that the location has been allocated for the given region.  The 
"build_lists" task is an example of an optimization done to reduce the number of "downregion" calls.
While a task could iterate over the elements of a subregion through repeated uses of "downregion" on
each element of the parent region, by building a list typed on the subregion up front, the cost 
of those "downregion" checks can be amortized over multiple iterations of the simulation.

Both reviewers A and B inquired how the dynamic disjointness tests are performed.  This
is a primary subject of our prior work [1].

Reviewer B commented that the Legion type system is similar in several ways to existing region-based
typed systems.  While we reuse many features of prior region languages, we believe that Legion
represents a fundamentally new composition of these features.  Primarily Legion allows for first-class
logical regions which permit elements to exist in many regions simultaneously via multiple dynamic
partitions of regions.  Furthermore, the way the region information is leveraged both at compile-time 
and at runtime represent a new point in the design space of region languages.  While the proof of
soundness does not introduce novel proof techniques, it is far from immediate that the type system
is sound and we felt the proof was necessary given the large differences in the semantics of logical regions 
from previous region systems.  The soundness proof was also a prerequisite for our later proofs
concerning coherence modes and hierarchical scheduling.

Both reviewers B and C thought coherence modes were an interesting technical contribution, but
would have liked a higher-level description to build intuition.
A description of the usage and intuition behind coherence modes is presented in [1] and
was omitted from this paper due to space constraints.  In this paper we present
two new results about coherence modes that are not in [1].  First, we present a formal semantics 
of the possible interleavings allowed by different coherence modes and their interactions.  
Second, we prove that coherence modes do not impact the soundness of our type system or the 
correctness of our hierarchical scheduling algorithm.  A brief summary of coherence modes is:
"exclusive" coherence corresponds to apparently-sequential execution, "atomic" corresponds 
to transactional-memory-like execution, and "simultaneous" corresponds to
"all bets are off" pthreads-style parallelism.  However, unlike other implementations, 
Legion's coherence modes only allow for reordering of subtasks with the same parent task, 
permitting the containment of non-determinism to a portion of the task hierarchy.

@@

Reviewer D asked what is currently implemented.  Legion currently consists of a static type 
checker for programs written in Core Legion and a C++ runtime library for fully-featured Legion programs.

Finally, reviewers B, C, and D expressed confusion about the nature of the clobber set.
The choice of large-step operational semantics was made to match the hierarchical and distributed
scheduling used by the Legion runtime.  When coherence modes other than exclusive are used, there may be
multiple tasks modifying the same location of memory - the clobber set is the set of all such locations for
a given task.  The clobber set is used to bound the allowed non-determinism within a given task, and the
fact that the top-level task (which has no sibling tasks) has an empty clobber set guarantees that any 
execution of the program is consistent with a standard sequentially consistent memory model.

An advantage of the large-step semantics with the use of clobber sets is that our soundness result holds for
any scheduling of tasks that satisfies the "valid_interleave" rules.  We considered the use of small-step semantics
that explicitly track the state of every parallel task's execution, but the complexity of the constraints used
to decide which task(s) were able to take a step at any given point were onerous and would limit our soundness
guarantees to only a particular scheduling algorithm.

Reviewer A asked why the E-NewColor rule allows the initial coloring to be non-empty.  There are two
reasons for this.  First, in the large-step semantics we use, the region mapping and heap typing cannot
be changed by an E-New rule, so a partitioning of a region must include extra elements of the appropriate
heap type to support allocation within a subregion.  Second, it allows implementations in which 
an existing element that was not colored by the programmer is included in (at most one) arbitrary
subregion, which can improve the efficiency of data transfers at runtime.

### 2012-09-12 doc/popl2013/response.txt

We thank the reviewers for their very detailed comments.  Most of the
comments are specifically about presentation; we acknowledge the
issue, and if the paper is accepted we will address the concerns
raised.

Many of the reviewers raised similar questions, which we have grouped
together in a single response below in addition to answering issues
raised by individual reviewers.  To keep the response to a reasonable
length we have tried to address the most common and significant
issues.  In many cases, questions were asked which were addressed in
our prior published work [1], which may have been unavailable to the
reviewers previously but can now be found at
http://theory.stanford.edu/~aiken/publications/papers/sc12.pdf .

Reviewers A, B, and C all asked for additional details about colorings which were covered in [1].
In Core Legion, colorings are explicit mappings of region elements (named by their location)
to integer "colors".  This represents the most general case of describing the partitioning of
a region.  More concise descriptions exist in the full Legion language (e.g. using a pure
function to determine the color of an element on demand), but a single form was selected for
Core Legion for simplicity.  As described in [1], Legion does not attempt to solve the (in many cases still-open)
problem of partitioning data for any given application.  In our example, the body of the "color_circuit"
task would require the programmer to compute a partition of the circuit graph and return a coloring
that assigns each circuit element a color corresponding to its piece of the partition.  The same
set of colors are used to identify the subregions that result from the partitioning.
In production scientific codes partitioning is often done by a call 
to an external library (e.g. PARMETIS for graph-based data structures).  
A coloring is merely a way to capture the result of the partitioning 
computation in a general format understood by the Legion runtime.

Reviewer C ask for clarification concerning the the relationship between multiple partitions and privileges.
The principle reason to support multiple partitions of the same data is to allow precise descriptions
of the sets of elements that will be accessed by a given task.  These sets are often different in 
different phases of an algorithm, and the increased precision is necessary to safely parallelize
applications such as the circuit example.  Although the subregions of different partitions are often
accessed with different privileges, this is a property of an algorithm rather than a Legion requirement.

### 2012-10-13 doc/types-pldi2013/langdef.tex
checkpoint

@@

Core Legion is explicitly typed using judgments of the form
$$\typeenv{e}{T}$$
where:
\begin{itemize}
\item $\Gamma$ is the type environment, mapping the names of local variables and global functions
to their types,
\item $\Phi$ is the region access privileges under which the expression is evaluated, and
\item $\Omega$ is the set of constraints that must hold between regions
\end{itemize}
As regions are hierarchical, privileges and/or constraints on a region often imply similar
privileges or constraints on subregions.  Figure~\ref{fig:closure} defines closure
operations for $\Phi^*$ and $\Omega^*$ that explicitly add these implied constraints.
The type checking rules are in Figure~\ref{fig:types}.  Many of the rules are 
straightforward, but several deserve more explanation.

Core Legion has no subtyping. A pointer's region is changed only via explicit
{\em upregion} and {\em downregion} operations.  {\em Upregion} calls can be inferred,
but {\em downregion} must be accompanied by a run-time check that the pointer really does
point into the subregion(s), so {\em upregion} is made explicit as well for simplicity.

Accesses to heap memory are performed via the {\em read}, {\em write}, and {\em reduce}
operations, which are only legal if the regions in a pointer's type are ones for which the
enclosing function has privileges.
Similarly, a call to another function is permitted only if the privileges required by the 
callee are a subset of those held by the caller.  There is no way to create privileges within
a Legion function - it is limited to privileges given to it by its caller.  (The
top-level task of a Core Legion program begins with a single, empty region from which other
regions can be partitioned and full read and
write privileges.)

The {\em partition} operation uses a {\em coloring} to split a region into 
disjoint subregions.  The subregions are given names within the body of the partition,
as well as constraints asserting subregions are pairwise disjoint and contained within the parent region.
%These constraints can be used locally, or packed into region relation instances that can be unpacked and used later.

The {\em pack} expression packs together a value and the corresponding logical regions,
provided the region relation's constraints hold.  The
corresponding {\em unpack} expression gives (new) names to those regions.  For simplicity, in Core Legion neither the 
{\em partition} nor the {\em unpack} expression allows shadowing of region names.

>>>>

As Legion attempts to benefit from both static and dynamic checks
to achieve performance while maintaining safety, both the type system and the operational
semantics have been extended in interesting ways from a basic expression language.  We
will describe these extensions in the next two subsections, and explain how these extensions
were chosen to preserve a straightforward proof of the soundness of the type system.

@@

Core Legion is explicitly typed using judgments of the form
$$\typeenv{e}{T}$$

In addition to the traditional type environment $\Gamma$, a Legion type judgment includes the
set of access privileges $\Phi$ for the logical regions in the expression as well as a set of
constraints $\Omega$ that must hold between those logical regions.

Both $\Phi$ and $\Omega$ are used in the heap access expressions: {\em read}, {\em write}, and {\em reduce}.
For a heap access to be valid, the the appropriate permission must exist for logical region(s) in the
pointer's type.  As regions are hierarchical, the exact region need not be named in $\Phi$ if permissions
exist for a logical region that is known to contain the pointer's region.  To simplify this check, we
define closure operations in Figure~\ref{fig:closure} that expand $\Omega$ into an $\Omega^*$, which is
then used to expand $\Phi$ into a $\Phi^*$ that does explicitly name every logical region for which
privileges are known to exist.

Core Legion provides no automatic pointer type conversions.  Valid pointers are created via the {\em new}
expression in specified region and invalid (i.e. {\em null}) pointers must also be typed.  Pointers can be
``upcast'' via the {\em upregion} expression, which uses the expanded constraints $\Omega^*$ to verify that
every possible region a pointer might point into is covered by at least one of the regions in the desired
pointer type.  In contrast, a ``downcast'' using the {\em downregion} expression must perform a dynamic
test to verify which region a run-time pointer value points into.  A static test similar to the one used
for {\em upregion} could be used to catch cases in which the downcast can never succeed at runtime, but we
have left it out for simplicity.

The introduction of constraints into $\Omega$ occurs through the use of the {\em partition} expression,
which makes use of a new type specific to Legion, a {\em coloring}.  A coloring is an opaque mapping from
pointers (which much point into the region being colored) to integer ``color'' values.  (For simplicity, 
Core Legion allows only an iterative construction of a coloring via a {\em newcolor} expression that generates
an empty map and a {\em color} expression which adds a new entry to an existing map, but the opacity of the
coloring allows for more efficient generation (and storage) of colorings as needed.)  The {\em partition} 
expression introduces names for subregions corresponding to each of those ``colors'', guaranteeing that each
subregion is disjoint from the other subregions and all are included in the original region.

The {\em pack} and {\em unpack} expressions allow the storage of pointers in the heap.  Logical region names
are lexically scoped and if stored in the heap as-is, can only be read within the same scope.  A region relationship
is used instead to capture the (dynamic) physical regions while preserving the statically known constraints
between those regions.  Because the {\em pack} expression is able to verify these constraints hold when the 
region relationship value is created, the {\em unpack} expression is able to reintroduce these constraints on
the (fresh) logical region names given to the contents of an unwrapped region relationship value.

Finally, the task call expression handles the aliasing of logical region names between the caller and callee and
verifies the property that the privileges held by a called task are always a subset of those held
by the caller.  In addition to being necessary for the proof of soundness, this property is critical to
enable the hierarchical scheduling model used by the Legion runtime.
